---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(tibble.print_min = 5, tibble.print_max = 5)
```

# miplicorn

<!-- badges: start -->
[![R-CMD-check](https://github.com/bailey-lab/miplicorn/workflows/R-CMD-check/badge.svg)](https://github.com/bailey-lab/miplicorn/actions)
[![Codecov test coverage](https://codecov.io/gh/bailey-lab/miplicorn/branch/master/graph/badge.svg)](https://codecov.io/gh/bailey-lab/miplicorn?branch=master)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

miplicorn establishes a unified framework for molecular inversion probe (MIP)
and amplicon analysis. It provides the tools to quickly parse files several
gigabytes large, filter and manipulate data, analyze data, and, most
informatively, visualize data.

## Installation
You may install the package from
[Github](https://github.com/bailey-lab/miplicorn) using `devtools`. Note that as
this is still a private repo, you must enter an authorization token to
`devtools::install_github`. Follow the instructions at [Managing Git(Hub)
Credentials](https://usethis.r-lib.org/articles/articles/git-credentials.html)
to generate a personal access token (PAT).
```{r installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github(repo = "https://github.com/bailey-lab/miplicorn",
                         auth_token = "<your PAT token>")
```

## Usage
See `vignette("miplicorn")` for a more extensive introduction and a
demonstration of several features of the package.
```{r usage}
library(miplicorn)

file <- miplicorn_example("reference_AA_table.csv")

data <- read_file(file, gene == "atp6")
data  

arrange_natural(data, sample, targeted)
```

