---
title: "FAQ"
description: "A compilation of frequently asked questions."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(miplicorn)
```

```{r example data, include = FALSE}
data <- tibble::tribble(
  ~sample, ~gene_id, ~gene, ~mutation_name, ~exonic_func, ~aa_change, ~targeted, ~coverage,
  "LA-05-37", "PF3D7_0709000", "crt", "crt-Cys72Ser", "missense_variant", "Cys72Ser", "Yes", 1716,
  "KO-05-62", "PF3D7_1133400", "PF3D7_1133400", "PF3D7_1133400-Glu187Lys", "missense_variant", "Glu187Lys", "No", 246,
  "HO-05-13", "PF3D7_1451200", "PF3D7-1451200", "PF3D7-1451200-Asn71Asn", "synonymous_variant", "Asn71Asn", "Yes", 1,
  "TO-05-43", "PF3D7_1133400", "PF3D7_1133400", "PF3D7_1133400-Glu121Lys", "missense_variant", "Glu121Lys", "No", 40,
  "KO-05-45", "PF3D7_1133400", "PF3D7_1133400", "PF3D7_1133400-Glu405Lys", "missense_variant", "Glu405Lys", "No", 0,
  "KS-05-95", "PF3D7_0709000", "crt", "crt-Ala220Ser", "missense_variant", "Ala220Ser", "Yes", 1,
  "AR-05-25", "PF3D7_0810800", "dhps", "dhps-Ile431Val", "missense_variant", "Ile431Val", "Yes", 274,
  "HO-05-52", "PF3D7_1133400", "PF3D7_1133400", "PF3D7_1133400-Ser283Leu", "missense_variant", "Ser283Leu", "No", 736,
  "KN-05-54", "PF3D7_1012700", "pph", "pph-Asn1189_Asn1194del", "disruptive_inframe_deletion", "Asn1189_Asn1194del", "No", 0,
  "LA-05-100", "PF3D7_0417200", "dhfr-ts", "dhfr-ts-Cys59Arg", "missense_variant", "Cys59Arg", "Yes", 17
)
```


## How can I add metadata to my data?

To add metadata to your metadata, you must read the metadata file as an R object
and then combine the result with your data.

Metadata is often saved as a delimited file—most commonly a CSV (comma separated
values) or a TSV (tab separated values)—or an excel file. There are several
packages that can be used to read in such files, but we recommend the use of
[`{vroom}`](https://vroom.r-lib.org/index.html) for delimited files and
[`{readxl}`](https://readxl.tidyverse.org/) for excel files.

```{r read metadata, eval = FALSE}
metadata <- vroom::vroom("path/to/file")
# OR
metadata <- readxl::read_excel("path/to/file")
```

```{r example metadata, include = FALSE}
metadata <- tibble::tribble(
  ~ID, ~District, ~Facility,
  "LA-05-37", "Lamwo", "Padibe",
  "KO-05-62", "Kole", "Aboke",
  "HO-05-13", "Hoima", "Kigorobya",
  "TO-05-43", "Tororo", "Nagongera",
  "KO-05-45", "Kole", "Aboke",
  "KS-05-95", "Kasese", "Karambi",
  "AR-05-25", "Arua", "Opia",
  "HO-05-52", "Hoima", "Kigorobya",
  "KN-05-54", "Kanungu", "Kihihi",
  "LA-05-100", "Lamwo", "Padibe"
)
```

After reading in your metadata, you can add the metadata to your data frame
using one of [`{dplyr}`'s mutating
joins](https://dplyr.tidyverse.org/reference/mutate-joins.html).

```{r combine data & metadata}
dplyr::left_join(data, metadata, by = c("sample" = "ID"))
```

Note that you may need to change the `by` argument to reflect the proper column
name of the metadata.

### My sample names do not match

In the case where your sample names do not match, first make sure that you are
reading in the correct files! In some cases, your data and metadata sample names
may be slightly different. For example, consider a dataset with the following
sample names:

```{r misformatted sample names, echo = FALSE}
data_misformatted <- tibble::tribble(
  ~sample, ~gene_id, ~gene, ~mutation_name, ~exonic_func, ~aa_change, ~targeted, ~coverage,
  "LA-05-37-ug-sur-2020-1", "PF3D7_0709000", "crt", "crt-Cys72Ser", "missense_variant", "Cys72Ser", "Yes", 1716,
  "KO-05-62-ug-sur-2020-1", "PF3D7_1133400", "PF3D7_1133400", "PF3D7_1133400-Glu187Lys", "missense_variant", "Glu187Lys", "No", 246,
  "HO-05-13-ug-sur-2020-1", "PF3D7_1451200", "PF3D7-1451200", "PF3D7-1451200-Asn71Asn", "synonymous_variant", "Asn71Asn", "Yes", 1,
  "TO-05-43-ug-sur-2020-1", "PF3D7_1133400", "PF3D7_1133400", "PF3D7_1133400-Glu121Lys", "missense_variant", "Glu121Lys", "No", 40,
  "KO-05-45-ug-sur-2020-1", "PF3D7_1133400", "PF3D7_1133400", "PF3D7_1133400-Glu405Lys", "missense_variant", "Glu405Lys", "No", 0,
  "KS-05-95-ug-sur-2020-1", "PF3D7_0709000", "crt", "crt-Ala220Ser", "missense_variant", "Ala220Ser", "Yes", 1,
  "AR-05-25-ug-sur-2020-1", "PF3D7_0810800", "dhps", "dhps-Ile431Val", "missense_variant", "Ile431Val", "Yes", 274,
  "HO-05-52-ug-sur-2020-1", "PF3D7_1133400", "PF3D7_1133400", "PF3D7_1133400-Ser283Leu", "missense_variant", "Ser283Leu", "No", 736,
  "KN-05-54-ug-sur-2020-1", "PF3D7_1012700", "pph", "pph-Asn1189_Asn1194del", "disruptive_inframe_deletion", "Asn1189_Asn1194del", "No", 0,
  "LA-05-100-ug-sur-2020-1", "PF3D7_0417200", "dhfr-ts", "dhfr-ts-Cys59Arg", "missense_variant", "Cys59Arg", "Yes", 17
)
data_misformatted$sample
```

In order to add your metadata to this dataset, you must first perform some
string manipulation so that the sample names in the dataset and metadata align.
To do so, you can leverage the [`{stringr}`](https://stringr.tidyverse.org/)
package.

```{r parse sample names}
data_formatted <- data_misformatted %>%
  dplyr::mutate(sample = stringr::str_remove(sample, "-ug.*"))
```

You can then use
[`{dplyr}`](https://dplyr.tidyverse.org/reference/mutate-joins.html) as before:

```{r merge formatted data with metadata}
dplyr::left_join(data_formatted, metadata, by = c("sample" = "ID"))
```

## How can I combine multiple probe sets?

In order to combine multiple probe sets, it is first important to ensure that
you plan to conduct the same analysis for each probe set. If you are not, you
may want to reconsider combining multiple probe sets. Assuming you want to
combine the same type of data for each probe set, you can simply bind the rows
using [`{dplyr}`](https://dplyr.tidyverse.org/reference/bind.html?q=bind).

```{r probes, include = FALSE}
probe_1 <- tibble::tribble(
  ~sample, ~chrom, ~pos, ~ref, ~alt, ~mutation_name, ~targeted, ~coverage,
  "AR-05-54", "chr13", "1629329", "AATAAATATATATAT", "A", "chr13:1629329:AATAAATATATATAT:A", "No", 0,
  "AR-05-73", "chr13", "1719135", "A", "T", "chr13:1719135:A:T", "No", 23,
  "KS-05-35", "chr8", "495991", "C", "T", "chr8:495991:C:T", "No", 0,
  "KN-05-43", "chr13", "1701344", "TAT", "A", "chr13:1701344:TAT:A", "No", 10,
  "MU-05-69", "chr13", "1713479", "C", "T", "chr13:1713479:C:T", "No", 3,
  "JI-05-82", "chr13", "1706967", "C", "A", "chr13:1706967:C:A", "No", 0,
  "JI-05-28", "chr13", "1706000", "A", "G", "chr13:1706000:A:G", "No", 10,
  "AG-05-28", "chr13", "1706978", "GATTAGATAAAAG", "TATTATAAAAAAAA", "chr13:1706978:GATTAGATAAAAG:TATTATAAAAAAAA", "No", 0,
  "AG-05-82", "chr4", "848676", "T", "A", "chr4:848676:T:A", "No", 0,
  "KS-05-63", "chr4", "736960", "C", "T", "chr4:736960:C:T", "No", 0
)

probe_2 <- tibble::tribble(
  ~sample, ~chrom, ~pos, ~ref, ~alt, ~mutation_name, ~targeted, ~coverage,
  "AR-05-54", "chr13", "958382", "T", "C", "chr13:958382:T:C", "No", 241,
  "AR-05-73", "chr13", "1726643", "A", "G", "chr13:1726643:A:G", "No", 185,
  "KS-05-35", "chr13", "1649722", "T", "C", "chr13:1649722:T:C", "No", 14,
  "KN-05-43", "chr13", "958465", "A", "G", "chr13:958465:A:G", "No", 517,
  "MU-05-69", "chr13", "958389", "C", "T", "chr13:958389:C:T", "No", 0,
  "JI-05-82", "chr13", "748341", "CATATAT", "C", "chr13:748341:CATATAT:C", "No", 0,
  "JI-05-28", "chr13", "1649445", "C", "A", "chr13:1649445:C:A", "No", 78,
  "AG-05-28", "chr8", "550212", "G", "A", "chr8:550212:G:A", "No", 150,
  "AG-05-82", "chr13", "1649307", "T", "A", "chr13:1649307:T:A", "No", 3,
  "KS-05-63", "chr13", "958443", "T", "C", "chr13:958443:T:C", "No", 2
)

probe_3 <- tibble::tribble(
  ~sample, ~chrom, ~pos, ~ref, ~alt, ~mutation_name, ~targeted, ~coverage,
  "AR-05-54", "chr4", "1121437", "G", "A", "chr4:1121437:G:A", "No", 3,
  "AR-05-73", "chr8", "714194", "T", "TATATATATATATATATAG", "chr8:714194:T:TATATATATATATATATAG", "No", 0,
  "KS-05-35", "chr13", "1868715", "A", "C", "chr13:1868715:A:C", "No", 3,
  "KN-05-43", "chr4", "516691", "A", "G", "chr4:516691:A:G", "No", 60,
  "MU-05-69", "chr13", "2512415", "C", "T", "chr13:2512415:C:T", "No", 87,
  "JI-05-82", "chr4", "835402", "T", "C", "chr4:835402:T:C", "No", 1,
  "JI-05-28", "chr8", "554378", "TA", "T", "chr8:554378:TA:T", "No", 43,
  "AG-05-28", "chr8", "501607", "AAAAATATATATAT", "A", "chr8:501607:AAAAATATATATAT:A", "No", 0,
  "AG-05-82", "chr4", "1089817", "T", "C", "chr4:1089817:T:C", "No", 83,
  "KS-05-63", "chr4", "390103", "TTAATAATAATAATAATAA", "T", "chr4:390103:TTAATAATAATAATAATAA:T", "No", 67
)
```

```{r bind rows}
dplyr::bind_rows(
  list(probe_1 = probe_1, probe_2 = probe_2, probe_3 = probe_3),
  .id = "probe_set"
)
```
