---
title: "Introduction to miplicorn"
description: >
  Learn how to get started with the basics of miplicorn.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to miplicorn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(miplicorn)
library(ggplot2)
```

## Parse

While MIPTools provides several .csv files that can be analyzed, parsing such
files is difficult because of their non-rectangular structure. As such,
attempting to read these files with default parameters fails.

```{r MIPTools output, message = FALSE, echo = FALSE}
cov_file <- miplicorn::miplicorn_example("coverage_AA_table.csv")
vroom::vroom(cov_file, col_names = FALSE, n_max = 10, col_select = 1:5)
```

As visible, there are six rows of metadata, which specify the gene ID, gene,
mutation name, etc. The remaining rows contain the data we are actually
interested in: the rows contain the samples and the columns contain the
positions we are interested in. The difficult part of reading in these files is
that the metadata must be extracted and treated differently from the data
itself.

miplicorn, therefore, provides two functions: `read_file()` and `read()` which
quickly read in such non-rectangular files. The functions generate a tibble
where each row represents a sample and a position. Thus, there will be multiple
entries for each unique sample.

```{r miplicorn parsing}
cov_file <- miplicorn::miplicorn_example("coverage_AA_table.csv")

data <- miplicorn::read_file(cov_file)
data
```

## Manipulate

### Amino acids
 
In some the user may want to convert amino acid abbreviations from the three to
one letters abbreviation, or vice versa for easier interpretation of data.

```{r amino acid abbreviation}
data %>% 
  dplyr::mutate(aa_change = convert_three(aa_change))
```

### Sort

In plotting data, it is useful to be able to control the order in which data
appears. While `dplyr::arrange()` provides the functionality to sort numeric or
character data, it lacks the ability to naturally sort alphanumeric vectors,
vectors containing both letters and numerics. Furthermore, the ordering of data
is not kept when fed into plotting functions. `arrange_natural()` attempts to
address these limitations.

```{r sort}
arrange_natural(data, sample, gene)
```

## Visualize

There are an almost limitless different ways to visualize a single set of data.
While `miplicorn` cannot address every method, it aims to simplify the creation
of key figures.

**N.B. these have not yet been incorporated into miplicorn, but as time goes**
**on more and more visualization methods will be added.**

### Average coverage
```{r visualize, echo = F, messsage = F}
# For the purposes of visualization, we select a small subset of genes
relevant_mutations <- c(
  "dhps", "k13", "atp6", "crt", "cytb",
  "mdr1", "mdr2", "dhfr-ts"
)

# Determine average coverage
data <- data %>%
  dplyr::filter(gene %in% relevant_mutations) %>%
  dplyr::group_by(gene, mutation_name, aa_change) %>%
  dplyr::summarise(value = mean(as.numeric(value)), .groups = "keep") %>%
  arrange_natural(gene, mutation_name, aa_change)

# Plot
ggplot(data, aes(x = aa_change, y = value, fill = gene)) +
  geom_col() +
  labs(
    x = "Amino Acid Change",
    y = "Average Coverage", 
    title = "Average Coverage by Mutation Site"
  ) +
  scale_fill_brewer(name = "Gene", palette = "PuBuGn") +
  theme_miplicorn() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


## "Validated" mutations
```{r mutation validation, echo = F, messsage = F}
# Create a fake list of validated mutations
set.seed(100)
valid_mutations <- tibble::tibble(
  mutation = unique(data$mutation_name),
  validated = sample(
    c("validated", "suspect"), 
    replace = TRUE, 
    size = length(unique(data$mutation_name)))
  )

# Combine the fake list of validated mutations with the previous tibble
validated <- data %>%
  dplyr::left_join(valid_mutations, by = c("mutation_name" = "mutation")) %>%
  dplyr::relocate(validated, .after = mutation_name)
```

```{r mutation validation plot, echo = F, warning = F}
validated %>%
  dplyr::group_by(mutation_name, aa_change, gene, validated) %>%
  dplyr::summarise(value = mean(value), .groups = "keep") %>%
ggplot(aes(x = aa_change, y = value, fill = gene, alpha = validated)) +
  geom_col() +
  labs(y = "Average Coverage", x = "Amino Acid Change", title = "Average Coverage by Mutation Site") +
  scale_fill_brewer(name = "Gene", palette = "PuBuGn") +
  theme_miplicorn() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

validated %>%
  dplyr::group_by(mutation_name, aa_change, gene, validated) %>%
  dplyr::summarise(value = mean(value), .groups = "keep") %>%
ggplot(aes(x = aa_change, y = value, fill = gene)) +
  geom_col() +
  facet_grid(validated ~ .) +
  labs(y = "Average Coverage", x = "Amino Acid Change", title = "Average Coverage by Mutation Site") +
  scale_fill_brewer(name = "Gene", palette = "PuBuGn") +
  theme_miplicorn() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```
